---
title: "Model Fitting Exercise – Mavoglurant"
author: "Alex Tejada-Strop"
format:
  html:
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
# Load required packages
library(tidyverse)
library(tidymodels)
library(here)
```

```{r}
# Load data
dat_raw <- read_csv(here("data", "Mavoglurant_A2121_nmpk.csv"))

glimpse(dat_raw)


dat_raw %>%
  ggplot(aes(x = TIME, y = DV, group = ID)) +
  geom_line(alpha = 0.3) +
  facet_wrap(~ DOSE) +
  labs(
    title = "DV over time by individual (raw data)",
    x = "Time",
    y = "DV"
  )

```

```{r}
dat_occ1 <- dat_raw %>%
  filter(OCC == 1)

dat_occ1 %>% count(OCC)


y_df <- dat_occ1 %>%
  filter(TIME != 0) %>%
  group_by(ID) %>%
  summarise(Y = sum(DV, na.rm = TRUE), .groups = "drop")

dim(y_df)

dose_df <- dat_occ1 %>%
  filter(TIME == 0)

dim(dose_df)

dat_joined <- dose_df %>%
  left_join(y_df, by = "ID")

dim(dat_joined)
glimpse(dat_joined)

dat_clean <- dat_joined %>%
  transmute(
    Y = Y,
    DOSE = DOSE,
    AGE = AGE,
    SEX = factor(SEX),
    WT = WT,
    HT = HT
  )

dim(dat_clean)
glimpse(dat_clean)
summary(dat_clean)

```

```{r}
dat_clean %>%
  summarise(
    n = n(),
    Y_mean = mean(Y, na.rm = TRUE),
    Y_sd = sd(Y, na.rm = TRUE),
    age_mean = mean(AGE, na.rm = TRUE),
    wt_mean = mean(WT, na.rm = TRUE),
    ht_mean = mean(HT, na.rm = TRUE)
  )

dat_clean %>%
  ggplot(aes(x = factor(DOSE), y = Y)) +
  geom_boxplot() +
  labs(
    title = "Total drug exposure (Y) by dose",
    x = "Dose",
    y = "Y"
  )

dat_clean %>%
  ggplot(aes(Y)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Y")

dat_clean %>%
  ggplot(aes(AGE)) +
  geom_histogram(bins = 20) +
  labs(title = "Distribution of AGE")

dat_clean %>%
  ggplot(aes(WT)) +
  geom_histogram(bins = 20) +
  labs(title = "Distribution of WT")

dat_clean %>%
  ggplot(aes(HT)) +
  geom_histogram(bins = 20) +
  labs(title = "Distribution of HT")
```

```{r}
rngseed <- 1234

set.seed(rngseed)
split_obj <- initial_split(dat_clean, prop = 0.75)

train_df <- training(split_obj)
test_df  <- testing(split_obj)

dim(train_df)
dim(test_df)

lm_spec <- linear_reg() %>%
  set_engine("lm")

# Model 1: DOSE only
lm_dose_fit <- lm_spec %>%
  fit(Y ~ DOSE, data = train_df)

# Model 2: all predictors
lm_all_fit <- lm_spec %>%
  fit(Y ~ DOSE + AGE + SEX + WT + HT, data = train_df)

lm_dose_preds_train <- predict(lm_dose_fit, new_data = train_df) %>%
  bind_cols(train_df %>% select(Y))

lm_all_preds_train <- predict(lm_all_fit, new_data = train_df) %>%
  bind_cols(train_df %>% select(Y))

rmse_dose <- rmse(lm_dose_preds_train, truth = Y, estimate = .pred)
rmse_all  <- rmse(lm_all_preds_train, truth = Y, estimate = .pred)

mean_y <- mean(train_df$Y)

null_preds <- train_df %>%
  mutate(.pred = mean_y)

rmse_null <- rmse(null_preds, truth = Y, estimate = .pred)

rmse_null
rmse_dose
rmse_all

rmse_compare_train <- bind_rows(
  rmse_null %>% mutate(model = "Null model"),
  rmse_dose %>% mutate(model = "Model 1: DOSE only"),
  rmse_all %>% mutate(model = "Model 2: All predictors")
) %>%
  select(model, .estimate)

rmse_compare_train

```

The null model had the highest RMSE, which means it performed the worst. The model using dose only improved prediction, and the full model with all predictors had the lowest RMSE, so it performed the best. The cross validation RMSE values were slightly higher than the training RMSE values, which is expected because cross validation evaluates model performance on unseen data. The overall ranking of the models stayed the same, suggesting the results were consistent.

```{r}
set.seed(rngseed)
folds <- vfold_cv(train_df, v = 10)

wf_dose <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE)

wf_all <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE + AGE + SEX + WT + HT)

cv_dose <- fit_resamples(
  wf_dose,
  resamples = folds,
  metrics = metric_set(rmse)
)

cv_all <- fit_resamples(
  wf_all,
  resamples = folds,
  metrics = metric_set(rmse)
)

cv_dose_metrics <- collect_metrics(cv_dose)
cv_all_metrics  <- collect_metrics(cv_all)

cv_dose_metrics
cv_all_metrics

cv_compare <- bind_rows(
  cv_dose_metrics %>% mutate(model = "Model 1: DOSE only"),
  cv_all_metrics %>% mutate(model = "Model 2: All predictors")
) %>%
  select(model, .metric, mean, n, std_err)

cv_compare

train_summary <- rmse_compare_train %>%
  rename(value = .estimate) %>%
  mutate(type = "Training RMSE")

cv_summary <- bind_rows(
  cv_dose_metrics %>%
    transmute(model = "Model 1: DOSE only", value = mean, type = "CV RMSE"),
  cv_all_metrics %>%
    transmute(model = "Model 2: All predictors", value = mean, type = "CV RMSE")
)

bind_rows(train_summary, cv_summary)
```

```{r}
set.seed(999)
folds_seed2 <- vfold_cv(train_df, v = 10)

cv_dose_seed2 <- fit_resamples(
  wf_dose,
  resamples = folds_seed2,
  metrics = metric_set(rmse)
)

cv_all_seed2 <- fit_resamples(
  wf_all,
  resamples = folds_seed2,
  metrics = metric_set(rmse)
)

collect_metrics(cv_dose_seed2)
collect_metrics(cv_all_seed2)
```

The null model had the highest RMSE, so it performed the worst. The model using dose only improved prediction compared to the null model. The full model with all predictors had the lowest RMSE, which means it performed the best on the training data.

The cross validation results showed the same overall pattern as the training RMSE results. Model 2 still performed better than Model 1, which suggests that adding AGE, SEX, WT, and HT improves prediction beyond dose alone. The CV RMSE values were fairly close to the training RMSE values, and the standard errors were moderate, suggesting that model performance was reasonably stable across folds.


```{r}
null_preds_plot <- train_df %>%
  transmute(
    Y,
    .pred = mean_y,
    model = "Null model"
  )

dose_preds_plot <- lm_dose_preds_train %>%
  mutate(model = "Model 1: DOSE only") %>%
  select(Y, .pred, model)

all_preds_plot <- lm_all_preds_train %>%
  mutate(model = "Model 2: All predictors") %>%
  select(Y, .pred, model)

pred_plot_df <- bind_rows(
  null_preds_plot,
  dose_preds_plot,
  all_preds_plot
)

ggplot(pred_plot_df, aes(x = Y, y = .pred, color = model)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    title = "Observed versus predicted values on training data",
    x = "Observed Y",
    y = "Predicted Y",
    color = "Model"
  )
```  


```{r}
resid_plot_df <- lm_all_preds_train %>%
  mutate(residual = .pred - Y)

ggplot(resid_plot_df, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_cartesian(ylim = c(-2500, 2500)) +
  labs(
    title = "Residuals versus predicted values for Model 2",
    x = "Predicted Y",
    y = "Residuals"
  )
```

```{r}
set.seed(rngseed)
dat_bs <- bootstraps(train_df, times = 100)

pred_bs <- matrix(NA, nrow = 100, ncol = nrow(train_df))

for (i in 1:100) {
  dat_sample <- analysis(dat_bs$splits[[i]])
  
  bs_fit <- lm_spec %>%
    fit(Y ~ DOSE + AGE + SEX + WT + HT, data = dat_sample)
  
  pred_bs[i, ] <- predict(bs_fit, new_data = train_df)$.pred
}

preds <- apply(pred_bs, 2, quantile, c(0.055, 0.5, 0.945)) %>%
  t() %>%
  as.data.frame()

colnames(preds) <- c("lower", "median", "upper")

boot_plot_df <- train_df %>%
  select(Y) %>%
  bind_cols(
    lm_all_preds_train %>% select(.pred),
    preds
  )

ggplot(boot_plot_df, aes(x = Y)) +
  geom_point(aes(y = .pred), alpha = 0.7) +
  geom_point(aes(y = median), shape = 1) +
  geom_errorbar(aes(ymin = lower, ymax = upper), alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    title = "Observed versus predicted values with bootstrap uncertainty",
    x = "Observed Y",
    y = "Predicted Y"
  )


```


The bootstrap plot adds uncertainty around the predictions from Model 2. Some observations have wider intervals than others, which suggests more uncertainty in those predictions. The model captures part of the overall pattern, but the spread around the diagonal line shows that prediction error still remains.

# Use of AI tools
AI tools were used to assist with debugging, clarifying error messages, and reviewing tidymodels syntax. The overall workflow design, data cleaning decisions, model specification, and interpretation of results were completed by me. AI support was used primarily as a learning and troubleshooting aid.


# Predict on the test data using Model 2 fit on training data

```{r}
lm_all_preds_test <- predict(lm_all_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(Y))
```

```{r}
train_plot_df <- lm_all_preds_train %>%
  mutate(dataset = "Training data") %>%
  select(Y, .pred, dataset)

test_plot_df <- lm_all_preds_test %>%
  mutate(dataset = "Test data") %>%
  select(Y, .pred, dataset)

final_plot_df <- bind_rows(train_plot_df, test_plot_df)

ggplot(final_plot_df, aes(x = Y, y = .pred, color = dataset, shape = dataset)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    title = "Observed versus predicted values for training and test data",
    x = "Observed Y",
    y = "Predicted Y",
    color = "Dataset",
    shape = "Dataset"
  )
```

# Test RMSE for Model 2
```{r}
rmse_test_all <- rmse(lm_all_preds_test, truth = Y, estimate = .pred)

rmse_test_all
```

# Overall model assessment table
```{r}
overall_summary <- tibble(
  model = c(
    "Null model",
    "Model 1: DOSE only",
    "Model 2: All predictors",
    "Model 2: Test data"
  ),
  rmse = c(
    rmse_null$.estimate,
    rmse_dose$.estimate,
    rmse_all$.estimate,
    rmse_test_all$.estimate
  )
)

overall_summary
```