---
title: "Model Fitting Exercise – Mavoglurant"
author: "Alex Tejada Strop"
format: html
execute:
  echo: true
  warning: false
  message: false
---

```{r}

# Load required packages
library(tidyverse)
library(tidymodels)
```

```{r}
# Load data
dat_raw <- readr::read_csv("Mavoglurant_A2121_nmpk.csv")

# Look at structure
glimpse(dat_raw)
```

```{r}
# Plot over TIME for each person, 
dat_raw %>%
  ggplot(aes(x = TIME, y = DV, group = ID)) +
  geom_line(alpha = 0.3) +
  facet_wrap(~ DOSE) +
  labs(
    title = "DV over time by individual (raw data)",
    x = "Time",
    y = "DV"
  )
```




## Keep only OCC

```{r}
dat_occ1 <- dat_raw %>%
  filter(OCC == 1)


# Confirm OCC values
dat_occ1 %>% count(OCC)
```

# I kept only rows where OCC equals 1 to avoid repeated results.

```{r}
# Sum DV across time points (exclude TIME == 0 dosing rows)
y_df <- dat_occ1 %>%
  filter(TIME != 0) %>%
  group_by(ID) %>%
  summarise(Y = sum(DV, na.rm = TRUE), .groups = "drop")

dim(y_df)  
```

```{r}
dose_df <- dat_occ1 %>%
  filter(TIME == 0)

dim(dose_df)  
```

```{r}
dat_joined <- dose_df %>%
  left_join(y_df, by = "ID")

dim(dat_joined)  
glimpse(dat_joined)
```

## Clean dataset

```{r}
dat_clean <- dat_joined %>%
  transmute(
    Y = Y,
    DOSE = DOSE,
    AGE = AGE,
    SEX = factor(SEX),
    RACE = factor(RACE),
    WT = WT,
    HT = HT
  )

glimpse(dat_clean)
summary(dat_clean)
```

```{r}
dim(dat_clean)
```


## Exploratory data analysis (EDA)
```{r}
dat_clean %>%
  summarise(
    n = n(),
    Y_mean = mean(Y, na.rm = TRUE),
    Y_sd = sd(Y, na.rm = TRUE),
    age_mean = mean(AGE, na.rm = TRUE),
    wt_mean = mean(WT, na.rm = TRUE),
    ht_mean = mean(HT, na.rm = TRUE)
  )
```

## Boxplot

```{r}
dat_clean %>%
  ggplot(aes(x = factor(DOSE), y = Y)) +
  geom_boxplot() +
  labs(title = "Total drug (Y) by dose", x = "Dose", y = "Y")
```


## Histogram

```{r}
dat_clean %>% ggplot(aes(Y)) + geom_histogram(bins = 30) + labs(title = "Y distribution")
dat_clean %>% ggplot(aes(AGE)) + geom_histogram(bins = 20) + labs(title = "AGE distribution")
dat_clean %>% ggplot(aes(WT)) + geom_histogram(bins = 20) + labs(title = "WT distribution")
dat_clean %>% ggplot(aes(HT)) + geom_histogram(bins = 20) + labs(title = "HT distribution")
```

## Train test split
```{r}
set.seed(123)
split_obj <- initial_split(dat_clean, prop = 0.80)
train_df <- training(split_obj)
test_df  <- testing(split_obj)
```

```{r}
dim(train_df)
dim(test_df)
```

## Regression
```{r}
# 1) Linear regression model
lm_spec <- linear_reg() %>%
  set_engine("lm")

# 2) Fit model
lm_dose_fit <- lm_spec %>%
  fit(Y ~ DOSE, data = train_df)

# 3) Predict
lm_dose_preds <- predict(lm_dose_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(Y))

# 4) RMSE and R-squared
metrics(lm_dose_preds, truth = Y, estimate = .pred) %>%
  filter(.metric %in% c("rmse", "rsq"))

# Fit model using all predictors
lm_all_fit <- lm_spec %>%
  fit(Y ~ DOSE + AGE + SEX + RACE + WT + HT, data = train_df)

# Predict on test data
lm_all_preds <- predict(lm_all_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(Y))

# Compute RMSE and R-squared
metrics(lm_all_preds, truth = Y, estimate = .pred) %>%
  filter(.metric %in% c("rmse", "rsq"))
```

```{r}
# Fit model using all predictors
lm_all_fit <- lm_spec %>%
  fit(Y ~ DOSE + AGE + SEX + RACE + WT + HT, data = train_df)

# Predict
lm_all_preds <- predict(lm_all_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(Y))

# RMSE and R-squared
metrics(lm_all_preds, truth = Y, estimate = .pred) %>%
  filter(.metric %in% c("rmse", "rsq"))
```

```{r}
levels(dat_clean$SEX)

```

## Model fitting
```{r}
# Specify logistic regression model
log_spec <- logistic_reg() %>%
  set_engine("glm")

# Fit logistic model using DOSE only
log_dose_fit <- log_spec %>%
  fit(SEX ~ DOSE, data = train_df)

# Predict classes
log_dose_class <- predict(log_dose_fit, new_data = test_df, type = "class")

# Predict probabilities
log_dose_prob <- predict(log_dose_fit, new_data = test_df, type = "prob")

# Combine results
log_dose_results <- bind_cols(
  test_df %>% select(SEX),
  log_dose_class,
  log_dose_prob
)

# Accuracy
accuracy(log_dose_results, truth = SEX , estimate = .pred_class)  

# Check column names before ROC
names(log_dose_results)

# Fit logistic model using all predictors
log_all_fit <- log_spec %>%
  fit(SEX ~ DOSE + AGE + RACE + WT + HT + Y, data = train_df)

# Predict classes
log_all_class <- predict(log_all_fit, new_data = test_df, type = "class")

# Predict probabilities
log_all_prob <- predict(log_all_fit, new_data = test_df, type = "prob")

# Combine results
log_all_results <- bind_cols(
  test_df %>% select(SEX),
  log_all_class,
  log_all_prob
)

# Accuracy
accuracy(log_all_results, truth = SEX, estimate = .pred_class)

# Check column names
names(log_all_results)
```

```{r}
names(log_dose_results)
```

```{r}
levels(train_df$SEX)
```

```{r}
roc_auc(log_dose_results, truth = SEX, .pred_2)


```

## ## Use of AI tools
AI tools were used to assist with debugging, clarifying error messages, and reviewing tidymodels syntax. The overall workflow design, data cleaning decisions, model specification, and interpretation of results were completed by me. AI support was used primarily as a learning and troubleshooting aid.